---
title: 'Reproducible Research: Peer Assessment 1'
author: "Todd B"
date: "01/17/2015"
output: html_document
keep_md: true
---

## Loading and preprocessing the data
Load the data (i.e. read.csv() )

```{r, echo=TRUE}
data_orig <- read.csv("~/classes/ReproData/RepData_PeerAssessment1/activity.csv")
```

Process/transform the data (if necessary) into a format suitable for analysis.  For this part of the assignment, the missing values in the dataset are ignored.
```{r, echo=TRUE}
data_nna <- subset(data_orig, !is.na(data_orig$steps))
```


## Histogram of the total number of steps taken each day
```{r, echo=TRUE}
data_nna_by_date <- aggregate(data_nna$steps, by=list(data_nna$date), FUN=sum)
names(data_nna_by_date) <- c("interv", "sum")

barplot(data_nna_by_date$sum, ylab = "Steps", xlab="Date")
```

## Average daily activity pattern:
Mean total number of steps taken per day:
```{r, echo=TRUE}
mean(data_nna_by_date$sum) 
```
Median total number of steps taken per day:
```{r, echo=TRUE}
median(data_nna_by_date$sum) 
```

## Time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo=TRUE}
data_nna_by_interv <- aggregate(data_nna$steps, by=list(data_nna$interval), FUN=sum)
num_days <- length(unique(data_nna$date))
data_nna_by_interv <- cbind(data_nna_by_interv, data_nna_by_interv[,2]/num_days)
names(data_nna_by_interv) <- c("interv", "sum", "day_av")

num_interv <- dim(data_nna_by_interv)[1]
interv_av <- sum(data_nna_by_interv$sum)/num_interv 

plot.ts(data_nna_by_interv[,1], data_nna_by_interv[,2], ylab = "Steps", xlab="Interval", type = "l") 
lines(data_nna_by_interv$interv, cbind(data_nna_by_interv, interv_av)[,4])
```

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

The maximum number of steps is:
```{r, echo=TRUE}
max_val <- max(data_nna_by_interv[,2])
max_val
```
Which is found at interval number:
```{r, echo=TRUE}
max_int <- data_nna_by_interv[which(data_nna_by_interv[,2]== max_val),1] 
max_int
```
Which is at index number:
```{r, echo=TRUE}
which(data_nna_by_interv[,1]== max_int)
```

## Imputing missing values
Total number of missing values in the dataset (i.e. the total number of rows with NAs):
```{r, echo=TRUE}
dim(data_orig)[1] - dim(data_nna)[1] 
```

Created a new dataset that is equal to the original dataset but with the missing data filled in.  Missing data is filled in with the mean step value for that interval:
```{r, echo=TRUE}
data_imp <- data_orig
for (i in 1:dim(data_orig)[1]) {
    if (is.na(data_orig[i,1])) {
        step_ndx <- which(data_nna_by_interv$interv == data_orig$interv[i])
        data_imp$steps[i] <- data_nna_by_interv$day_av[step_ndx]
        #print(c(step_ndx, data_nna_by_interv$day_av[step_ndx]))
    }
}
```


## Histogram of the total number of steps taken each day. 
```{r, echo=TRUE}
data_imp_by_date <- aggregate(data_imp$steps, by=list(data_imp$date), FUN=sum)

data_imp_by_interv <- aggregate(data_imp$steps, by=list(data_imp$interval), FUN=sum)

barplot(data_imp_by_date[,2], ylab = "Steps", xlab="Date")
```

Mean and median total number of steps taken per day:
```{r, echo=TRUE}
mean(data_imp_by_date[,2]) 
median(data_imp_by_date[,2]) 
```


## Differences in activity patterns between weekdays and weekends:

Created a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r, echo=TRUE}
data_imp <- cbind(data_imp, as.factor((weekdays(strptime(data_imp[,2], format = "%Y-%m-%d")) == "Saturday" | weekdays(strptime(data_imp[,2], format = "%Y-%m-%d")) == "Sunday" )))
colnames(data_imp) <- c("steps", "date", "interval", "weekend")
```

Panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r, echo=TRUE}
data_tmp <- subset(data_imp, data_imp$weekend == FALSE)
data_imp_by_interv <- cbind(data_imp_by_interv, aggregate(data_tmp$steps, by=list(data_tmp$interval), FUN=sum)[2])

data_tmp <- subset(data_imp, data_imp$weekend == TRUE)
data_imp_by_interv <- cbind(data_imp_by_interv, aggregate(data_tmp$steps, by=list(data_tmp$interval), FUN=sum)[2])

names(data_imp_by_interv) <- c("interval", "steps", "wkday", "wkend")
sum(data_imp_by_interv$wkday)+sum(data_imp_by_interv$wkend) 

library(lattice)
par(mfrow=c(2,1))  
plot(data_imp_by_interv$interval, data_imp_by_interv$wkday, type="l", col="darkgrey",
xlab="Interval", ylab="Steps")
plot(data_imp_by_interv$interval, data_imp_by_interv$wkend, type="l", col = "brown", xlab="Interval", ylab="Steps" ) 
dev.off()
```
Comparison of the two plots reveal that there is a great deal of difference in activity between the weekdays and the weekend.  Mush more walking is done on the weekend.





